---
title: Using Survey Weight
author: Fahim Ahmad
date: '2022-05-14'
slug: using-survey-weight-in-r
categories: [R]
tags:
  - R
Description: In this post, you will learn how to use weights in R using the survey package.
output:
  blogdown::html_page: 
    toc: yes

Summary: In R working with survey weight is made possible using the `survey` package.
---

In R working with survey weight is made possible using `survey` package. Let's use below data frame as an exmaple here:

```{r}
set.seed(1000)
age <- c(18:100)
age <- sample(age, 100, replace = TRUE)
gender <- c("Male", "Female")
gender <- sample(gender, 100, replace = TRUE)
country <- c("A", "B")
country <- sample(country, 100, replace = TRUE)

df <- data.frame(age, gender, country)

df$weight[df$gender=="Female"] <-50/sum(df$gender=="Female")
df$weight[df$gender=="Male"] <-50/sum(df$gender=="Male")

# summary of data
summary(df)
```

The most important variable here is the _weight_ variable which is constructed to balance the sex ratio. 

Inside the `survey` package, there is `svydesign()` function that can be used to link a data frame with a weight.

```{r message=FALSE}
# install.packages("survey")
library(survey)
df.w <- svydesign(ids = ~1, data = df, weights = ~weight)
```

The resulting object is not a data frame anymore, but is a list of different objects that can be seen using `attributes()` function.

```{r}
attributes(df.w)
```

Therefore, we need to use `survey`'s own analytical functions. For example, here is a comparison of unweighted and weighted sex ratio.


```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(dplyr)
```

```{r}
# unweighted
df %>%
{table(.$gender)} %>%
  prop.table()

# weighted
df.w %>%
  svytable(~gender, .) %>%
  prop.table()
```

`svytable()` can be used to create more than one-way frequency/percentage tables as well. For example, let's create contingency table of `gender` and `country`

```{r}
df.w %>%
svytable(~gender+country, .) %>%
  prop.table(2)
```


Below are other useful functions of `survey` package:

```{r eval=FALSE, include=TRUE}
# to compute weighted mean
svymean(~age, df.w)
# to compute weighted quantiles
svyquantile(~age, df.w, c(.25, .50, .75))
# to compute weigted variance
svyvar(~age, df.w)
# to perform t-test
svyttest(age~gender, df.w)

```






