---
title: Using dplyr::across( ) with Anonymous Function
author: Fahim Ahmad
date: '2021-07-20'
slug: using-across-with-anonymous-function
categories:
  - R
tags:
  - R

output:
  html_document:
    keep_md: false
    toc: true
    self_contained: false
    theme: null
    template: "../src/blog-template.html"
---

In my previous [post](../2021-05-15-advantages-of-dplyr-across-to-apply-a-function-across-multiple-columns/), I briefly described the advantages of using `across()` while applying the same function over multiple columns. In this post, I provide an overview of using `across()` with the anonymous function.

I simulate the below data frame to use as an example here.

```{r}
set.seed(1000)
n <- 10
df <- data.frame(
  age = c(sample(18:30, n, replace = T), 80),
  income = c(sample(2000:3000, n, replace = T), 8000),
  education_level = c(sample(12:20, n, replace = T), 25),
  gender = c(sample(c("Male", "Female"), n, replace = T), "Male")
)
df
```

For instance, let’s say you decide to replace the outliers in numeric columns with the mean value before carrying out some statistical analysis.

>**Note**: The presence of outliers can be checked using the `boxplot.stats()` function from **grDevices** package. Use **?boxplot.stats** for more information.

First, let's replace the outliers with the mean value on one column at a time, and then the next column, and so on.

```{r message=FALSE, warning=FALSE}
library(dplyr)

df %>%
  mutate(
    age = case_when(
      age %in% boxplot.stats(age)$out ~ mean(age, na.rm = T),
      TRUE ~ age
    ),
    income = case_when(
      income %in% boxplot.stats(income)$out ~ mean(income, na.rm = T),
      TRUE ~ income
    ),
    education_level = case_when(
      education_level %in% boxplot.stats(education_level)$out ~ mean(education_level, na.rm = T),
      TRUE ~ education_level
    )
  )
```

Imagine what if you had 100 additional numeric columns in the data set from which you wanted to handle the outliers? using the above approach you would have to write a very long chunk of code inside the `mutate()` function.

Alternatively, you can create a custom function to deal with the outliers and apply it to all numeric columns using the `across()` function as below.

```{r}
# creating a function to replace outliers with the mean value
outliers_to_mean <- function(x) {
  x = case_when(
    x %in% boxplot.stats(x)$out ~ mean(x, na.rm = T),
    TRUE ~ x
  )
}

# using the custom function inside the across() function
df %>% 
  mutate(
    across(
      .cols = where(is.numeric),
      .fns = outliers_to_mean
    )
  )
```

The question is what if you need the `outliers_to_mean()` function only in this specific instance, and you know you would not use it again in the future?

In this case, it is better to use an anonymous function instead of a named function. For instance, here is how you can use the above code as an anonymous function that produces the same result.

```{r}
df %>% 
  mutate(
    across(
      .cols = where(is.numeric),
      .fns = function(x) {
        x = case_when(
          x %in% boxplot.stats(x)$out ~ mean(x, na.rm = T),
          TRUE ~ x
          )
      }
    )
  )

```


There is another way to even reduce the number of unnecessary stuff and write fewer lines of code. The `.fns` argument accepts purrr-style lambda as well. For example, in the above chunk of code, you can replace `function(x){}` with `~{}` which is just another way of saying that you have started an anonymous function.

>**Note**: while using `~`, the `.x` symbol can be used to refer to the argument of the anonymous function.

```{r}
df %>% 
  mutate(
    across(
      .cols = where(is.numeric),
      .fns = 
        ~{.x = case_when(
          .x %in% boxplot.stats(.x)$out ~ mean(.x, na.rm = T),
          TRUE ~ .x
          )
        }
    )
  )
```


