---
title: Rename Data Frame Columns
author: ''
date: '2019-06-19'
slug: rename-data-frame-columns
categories:
  - R
tags:
  - R
output:
  blogdown::html_page: 
    toc: yes
    
summary: In this post you will learn how to rename columns of a data frame in R.    
---

There are several ways of renaming variables in R. In this post you will learn how to change a column name using base R function, the dplyr way and using data.table package.

I will use the R built-in **iris** data frame.

```{r}
head(iris)
```

# Rename column names with R base functions

Let's change **Sepal.Length** to **S.P**. The procedure is simple; get column names using `names()` or `colnames`, extract the specific column name, and assign a new name.

```{r}
colnames(iris)[colnames(iris)=="Sepal.Length"] <- "S.P"
head(iris)
```

The code does the following:

1. `colnames(iris)` looks into all the names in the **iris**.
2. `names(iris) == "Sepal.Length"` returns a vector with true and false values to extract the variable name you want to change.
3. `<- "S.P"` assigns the new name.

You can also rename a variable by its index as well. For example, let's change **Sepal.Width** [the second column] to **S.W**.

```{r}
colnames(iris)[2] <- "S.W"
head(iris)
```

# Rename column names with dplyr

You can use `rename()` function to change column names as following. For instance, let's change **Petal.Length** to **P.L** and **Petal.Width** to **P.W**.

```{r message=FALSE}
# install.packages("dplyr")
library(dplyr)

iris %>%
  rename(
    P.L = Petal.Length,
    P.W = Petal.Width
  ) %>%
  head()
```

# Rename column names with data.table

`setnames()` function from **data.table** package can be used to rename a variable. The syntax is `setnames(df, "old_name", "new_name")`. As an example, let's change **Species** to **SPECIES**.

```{r include=FALSE}
rm(iris)
```


```{r message=FALSE}
# install.packages("data.table")
library(data.table)
setnames(iris,
         "Species", "SPECIES")
colnames(iris)
```


