---
title: 'ggplot2: geom_histogram & facet_wrap with different vertical lines on each facet'
author: Fahim Ahmad
date: '2022-04-03'
slug: ggplot2-facet-wrap-with-different-vertical-lines-on-each-facet
categories:
  - R
tags:
  - ggplot
  - visualization
Description: ""

output:
  html_document:
    keep_md: false
    toc: true
    self_contained: false
    theme: null
    template: "../src/blog-template.html"
---


The `geom_histogram()` function from **ggplot2** package is used to create a histogram plot. For example, let's plot the distribution of `Sepal.Length` from **iris** data.

```{r message=FALSE}
library(ggplot2)
theme_set(theme_bw())
ggplot(iris, aes(Sepal.Length)) + geom_histogram(fill = "orange")
```

To add a vertical line to show the mean value of `Sepal.Length`, we can use `geom_vline()`.

```{r message=FALSE}
library(ggplot2)
ggplot(iris, aes(Sepal.Length)) + geom_histogram(fill = "orange") +
  geom_vline(data = iris, aes(xintercept = mean(Sepal.Length)))
```


One of the most powerful aspects of **ggplot2** is the ease with which you can create multiple sub-plots using `facet_wrap()`. For example, let's plot the distribution of `Sepal.Length` by each group of `Species` from **iris** data.

```{r message=FALSE}
ggplot(iris, aes(Sepal.Length)) + geom_histogram(fill = "orange") +
  facet_wrap(~Species) +
  geom_vline(data = iris, aes(xintercept = mean(Sepal.Length)))
  
```

While using `facet_wrap()` each plot shows a different subset of the data, however, the `geom_vline()` adds the vertical line on the same x-intercept in all plots. For instance, in the above histograms the vertical line shows the overall mean of `Sepal.Length` column instead of mean of each group.

It's more practical to add a vertical line on each facet showing the mean for each group of data. To do so, a separate data frame containing the mean of each group should be created to use with `geom_vline()`.

```{r message=FALSE}
library(dplyr)
vline <- summarise(group_by(iris,Species), mean = mean(Sepal.Length))
vline

ggplot(iris, aes(Sepal.Length)) + geom_histogram(fill = "orange") +
  facet_wrap(~Species) +
  geom_vline(data = vline, aes(xintercept = mean))

```

