---
title: Visualizing COVID-19 Data in R | Johns Hopkins University
author: Fahim Ahmad
date: '2020-05-29'
slug: visualizing-covid-19-data-in-r-johns-hopkins-university
categories:
  - R
  - visualization
tags:
  - R
  - visualization

output:
  html_document:
    keep_md: false
    toc: true
    self_contained: false
    theme: null
    template: "../src/blog-template.html"
---
In this post, I obtain COVID-19 data from the Johns Hopkins University's GitHub account and explore how to visualize it specific to each country - Afghanistan as an example here.

First thing first, below packages should be installed and loaded.

```{r message=FALSE}
# install.packages("dplyr")
library(dplyr)
# install.packages("ggplot2")
library(ggplot2)
# install.packages("tidyr")
library(tidyr)
# install.packages("data.table")
library(data.table)
# install.packages("lubridate")
library(lubridate)
# install.packages("patchwork")
library(patchwork)
```

## Step 1: Load the COVID-19 confirmed, recovered, and deaths data

I am using the `fread()` function from **data.table** package for this purpose.

```{r eval=F}
covid19_confirmed_cases <- fread("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")

covid19_recovered <- fread("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_recovered_global.csv")

covid19_deaths <- fread("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")

```

## Step 2: Combine the data frames
Since there is a separate data frame for each type of COVID-19 cases, I combine all data frames using `rbind()` function.

```{r message=FALSE}
covid19 <- rbind(
  covid19_confirmed_cases %>% 
  mutate(type = "confirmed"),
  
  covid19_recovered %>% 
    mutate(type = "deaths"),
  
  covid19_deaths %>% 
    mutate(type = "recovered")
)
```

Let's have a look at the spread of COVID-19 around the globe.

```{r fig.height=10, fig.width=10, warning=FALSE}
covid19 %>%
  select(Lat, Long, value = colnames(covid19)[length(colnames(covid19))-1], type) %>% 
  ggplot() +
  geom_map(data = map_data('world'), map = map_data('world'), aes(x = long, y = lat, group = group, map_id=region), fill = "black", colour = "gray", size=0.5) +
  geom_point(aes(x = Long, y=Lat, size = value, color = type)) +
  facet_wrap(~type, ncol = 1) +
  scale_y_continuous(breaks=c(), limits = c(-60, 90)) +
  scale_x_continuous(breaks=c()) +
  labs(x=NULL, y=NULL) +
  theme(legend.title = element_blank(), legend.position = "top") +
  scale_color_manual(values = c("orange", "red", "green")) 
```

# Step 3: Subset the merged data frame
Using the `filter()` function from **dplyr** package you can filter the data for any specific country - Afghanistan as an example here^[For those countries that the data is recorded at the province/state level, sum of all provinces/states should be used as country level data.].

```{r}
covid19_AFG <- covid19 %>% 
  # filter the data where Country/Region is Afghanistan
  filter(`Country/Region` %in% "Afghanistan") %>% 
  # remove the Province/State, Country/Region, Lat, and Long columns
  select(-`Province/State`, -`Country/Region`, -Lat, -Long) %>% 
  # transform the data from wide to long format
  pivot_longer(-type, names_to = "date", values_to = "cases")
  
```

# Step 4: Plot the data
At this stage, let's have a glance at the structure of the data.

```{r}
str(covid19_AFG)
```

As seen above, the date column is stored as character format. One important point to always remember is while visualizing the time-series data, the time variable must be recorded as date/time format. To do so, I am using the `mdy()`^[I am using the mdy() function becuase the date is stored as Month/Day/Year. Other possible options are  dmy(), myd(), ymd(), ydm(), dym()...] function from **lubridate** package to transform the date column into a standard date format.

```{r}
covid19_AFG <- covid19_AFG %>% 
  mutate(date = mdy(date)) %>% 
  # optionally filter the data for any date range
  filter(date < ymd("2021-06-30"))
```

Next, I use the `geom_line()` function from **ggplot2** package to plot the data.

```{r}
plot <- covid19_AFG %>% 
  # initialize a ggplot object
  ggplot() +
  # specify the line plot aesthetics
  geom_line(aes(x = date, y = cases, color = type)) +
  # use the black and white theme and set the text angle in the x-axis to -45
  theme_bw() + theme(axis.text.x = element_text(angle = -45)) +
  # remove the x-axis, y-axis, and legend titles; add the plot title
  labs(x = NULL, y = NULL, color = NULL, title = "Cumulative COVID-19 Cases in Afghanistan")

plot
```

### Formatting the x-axis

The `scale_x_date()` function can be used to customize the date format in the x-axis. Below are some examples of how to use the function.

```{r fig.height=10, fig.width=10, message=FALSE}
# date in 'MM DD YY' Format. Months in abbreviated format
p1 <- plot + scale_x_date(date_labels = "%b %d - %y")

# date in 'MM DD YY' Format. Months in unabbreviated format
p2 <- plot + scale_x_date(date_labels = "%B %d - %Y")

# date in 'YYYY-MM-DD' format
p3 <- plot + scale_x_date(date_labels = "%Y-%m-%d")

(p1) / (p2 | p3) +
  plot_layout(guides = 'collect') &
  theme(legend.position = "top")
```

Here are the available options for customizing the date variable:

```{r echo=FALSE}
data.frame( Symbol = c("%d", "%a", "%A", "%m", "%b", "%B", "%y", "%Y"),
            Meaning = c("Days as number", "Abbreviated weekday", "Unabbreviated weekday", "Month as number", "Abbreviated month", "Unabbreviated month", "2-digit year", "4-digit year" ),
            Example = c("01-31", "Mon", "Monday", "01-12", "Jan", "January", "02", "2020")
) %>% 
  knitr::kable(caption = "List of available options for customizing the date variable.")
```

It is also possible to control the numbers of major breaks and minor breaks using the `date_breaks()` and `date_minor_breaks()` functions.

```{r fig.height=8, fig.width=10}
plot +
  # set the date format to Month+Day
  scale_x_date(date_labels = "%b %d",
               # set the date breaks to 1 month
               date_breaks = "1 month",
               # set the minor ticks in the x-axis to 1 week
               date_minor_breaks = "1 week")
```

The above plot shows the cumulative number of Covid19 cases in Afghanistan. Let’s compute the daily cases by deducting the value of each row from its previous row.

```{r}
covid19_AFG_daily_cases <- covid19_AFG %>% 
  group_by(type) %>% 
  mutate(cases = ifelse(row_number()==1, cases, cases - lag(cases, default = first(cases))))

covid19_AFG_daily_cases %>% 
  ggplot() +
  geom_line(aes(x = date, y = cases, color = type)) +
  labs(x = NULL, y = NULL, color = NULL, title = "Daily COVID-19 Cases in Afghanistan") +
  theme_bw() + theme(axis.text.x = element_text(angle = -45)) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 month", date_minor_breaks = "1 week")
```


